Making ice sheet models \\ scale properly

### 2023-01-23 talk/slides.tex
finally close to done with the draft talk ... but not done

\bigskip
\begin{itemize}
\item recall traditional numerical analysis applied to these equations
\end{itemize}

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{\item[]}
\only<2>{
\item we expect \alert{explicit} time stepping for advections
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] conditional stability
    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
    \end{itemize}
}
\only<3>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] or: Crank-Nicolson, BDF, \dots
    \item[$\circ$] unconditional stability
    \item[$\circ$] must solve equations at each step
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

@@

\begin{itemize}
\item current researchers mostly believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: the problem is a DAE
    \item[$\circ$] performance and design consequences: stability is determined by the coupled solution velocity $\bU$
    \end{itemize}

\medskip
\item<2> \alert{implicit time-stepping} is expected for DAE problems
    \begin{itemize}
    \item[$\circ$] the obvious alternative
    \end{itemize}
\item<2> a sequence of problems (``equations'') must be solved at each time step: NCP coupled to Stokes
\end{itemize}
\end{frame}

@@

\begin{frame}{analyzing ice sheet model performance}

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
\item $m$ = (\# of surface elevation, velocity, and pressure unknowns)
\item domain $\Omega \subset \RR^2$ (map-plane) has characteristic width $L$
\item map-plane (2D) mesh cells have characteristic diameter $h$:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}}$$
\item recall mass-continuity equation schemata:
{\footnotesize
\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}
\end{itemize}
\end{frame}

\begin{frame}{ice sheet model performance}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^2}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$
\end{tabular}
\end{frame}

@@

\begin{itemize}
\item main point: \alert{explicit time-stepping stability restrictions imply many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] $\alpha$ for stress balance solver important
    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix low model performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
        \begin{itemize}
        \item[$\vartriangleright$] or SIA NCP
        \item[$\vartriangleright$] in any case, a \alert{free-boundary problem}
        \item[$\vartriangleright$] next slide parameterizes expense of these coupled, implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
        \end{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \end{itemize}

\bigskip
\item<2> \alert{existing} implicit time-stepping ice sheet models:
    \begin{itemize}
    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
    \item[$\circ$] no other fully-implicit NCP solvers known?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{ice sheet model performance 2}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit \, {\footnotesize \emph{advective}} & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} \, {\footnotesize \emph{diffusive}}  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}
\end{frame}

@@

\begin{frame}{semi-coupled time-stepping}

@@

\begin{frame}{machine learning}

FIXME Jouvet et al.~2021

\begin{center}
\includegraphics[width=0.45\textwidth]{images/cnn-jouvet.png}

\includegraphics[width=0.65\textwidth]{images/tiles-jouvet.png}
\end{center}
\end{frame}

\begin{frame}{scalable obstacle problem solvers}

@@

\begin{frame}{scalable obstacle problem solvers}

>>>>

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
\item nonetheless, let us recall some traditional numerical analysis of these equations \dots
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
    \end{itemize}
}
\only<4>{
\item \alert{implicit} time stepping for \alert{diffusions} is recommended
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] unconditionally stable, but must solve equations at each step
    \item[$\circ$] alternative implicit schemes: Crank-Nicolson, BDF, \dots
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

@@

\begin{itemize}
\item current researchers mostly believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: \alert{the mathematical problem is a DAE}
    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$,
    \item[$\circ$] but there are performance and design consequences of hidden DAE \& diffusive character
    \end{itemize}

\medskip
\item<2> \alert{implicit time-stepping} is expected for DAE problems,
\item<2> but a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes at each time step}
    \begin{itemize}
    \item[$\circ$] not really ``equations'' any more because of $s\ge b$ inequality
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{simplified analysis of ice sheet model performance}

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
    \begin{itemize}
    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
    \end{itemize}
\item assume map-plane domain $\Omega \subset \RR^2$ of width $L$
\item assume map-plane mesh cells have characteristic diameter $h$
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item the question addressed by upcoming performance table:

\bigskip
\begin{center}
\begin{minipage}{0.82\textwidth}
how does flops per model year \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
\end{minipage}
\end{center}
\end{itemize}
\end{frame}

\begin{frame}{preparation: observe and recall}

\begin{itemize}
\item ice sheets are thin layers, thus even non-shallow Stokes models often have $O(1)$ mesh points in the vertical direction
    \begin{itemize}
    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
    \end{itemize}

\medskip
\item scale relationships:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$

\vspace{-18mm}
\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}

\medskip
\item mass-continuity equation time-stepping schemata:
{\footnotesize
\begin{align*}
\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
\end{align*}
}

\item stress-balance solver scaling parameterized as $O(m^{1+\alpha})$
\end{itemize}
\end{frame}

\begin{frame}{\emph{explicit} ice sheet model performance}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}

\bigskip\bigskip
\begin{itemize}
\item but what about implicit?
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item main point so far: \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] yes, stress balance solver scaling ($\alpha$) is important
    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
        \begin{itemize}
        \item[$\vartriangleright$] a \alert{free-boundary problem}
        \item[$\vartriangleright$] table on next slide parameterizes cost of coupled implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
        \end{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \end{itemize}

\bigskip
\item<2> \alert{existing} implicit time-stepping ice sheet models?
    \begin{itemize}
    \item[$\circ$] no fully-implicit NCP-coupled-to-Stokes models exist (?)
    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{ice sheet model performance table (Bueler, 2022)}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}
\end{frame}

@@

\begin{frame}{scalable approach 1: machine learning}

\begin{itemize}
\item apply \alert{machine learning}
\item \emph{concept (Jouvet et al.~2021)}: run existing non-scalable ice sheet models on a range of hypothetical/real ice sheets, and train an ML on the results
    \begin{itemize}
    \item[$\circ$] supervised learning of physically-based model yields \alert{emulator}
    \end{itemize}
\end{itemize}

\vspace{3mm}
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\only<1>{\includegraphics[height=0.52\textheight]{images/tiles-jouvet.png}}
\only<2>{\includegraphics[height=0.6\textheight]{images/cnn-jouvet.png}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{scalable approach 2: smarter explicit time-stepping}

semi-coupled

@@

\begin{frame}{scalable approach 3: multigrid implicit solvers}

@@

\begin{frame}{scalable approach 3: multigrid implicit solvers}
